Maximizing Ink in Symmetric Partial Edge Drawings of $k$-plane Graphs

### 2018-02-14 eurocg/paper.tex
minor remarks addressed

The idea of drawing edges only partially has been formalized in graph drawing as follows~\cite{bk-ecbe-12}. 
A \emph{partial edge drawing (PED)} is a graph drawing that maps vertices to points and edges to pairs of crossing-free edge stubs of positive length pointing towards each other.
These edge stubs are obtained by erasing one contiguous central piece of the straight-line segment connecting the two endpoints of each edge.
In other words each straight-line edge is divided into three parts, of which only the two outer ones are drawn (see Fig.~\ref{fig:examples}).
More restricted and better readable~\cite{blmt-pedhmitc-16} variations of PEDs are \emph{symmetric} PEDs, in which both stubs of an edge must have the same length (see Fig.~\ref{fig:examples}(b)), and \emph{homogeneous} PEDs, in which the ratio of the stub length to the total edge length is the same for all edges.
The natural optimization problem in this formal setting is \emph{ink maximization}, i.e., maximizing the total stub length, so that as much information as possible is given in the drawing while all crossings disappear in the negative background space.

@@

Throughout the paper let $ G $ be a \emph{simple graph} with edge set $ S = \{s_1,\dots,s_m\}$ and $ \Gamma $ a straight line drawing of $ G $ in the plane. We call $ \Gamma $ \emph{$ k $-plane} if every edge $ s_i \in S $ is crossed by at most $ k $ other edges from $ S $ in $ \Gamma $. We use the terms edge in $ S $ and segment in $ \Gamma $ interchangeably. %, since our input is always a graph $G$ with a given drawing $ \Gamma $. 
Hence we can also interpret $S$ as a set of line segments.

@@

%\renewcommand\windowpagestuff{%
%	\includegraphics{Reduction_visual}
%	\captionof{figure}{Example of three variable gadgets and\\ a satisfied clause gadget. Dotted parts do not belong\\ to the SPED.}
%	\label{fig:gadget_connection}
%}
%\opencutleft
\begin{proof}
%		\begin{cutout}{2}{0pt}{0.44\textwidth}{15}
	We reduce from the \NP-hard problem \ppsat~\cite{l-pftu-82} using similar ideas as in Bruckdorfer's general proof of the hardness of \maxsped~\cite{b-sgh-15}. %, which reduces from the more restricted problem \textsc{Planar Positive 1-in-3 Sat}. 
	Let $\phi$ be a planar 3-Sat formula with $n$ variables $\{x_1, \dots, x_n\}$ and $m$ clauses $\{c_1, \dots, c_m\}$, each consisting of three literals.
	We can assume that $\phi$ comes with a planar drawing of its variable-clause graph $H_\phi$, which has a vertex for each variable $x_i$ and a vertex for each clause $c_j$.
	Each clause vertex is connected to the three variables appearing in the clause.
	In the drawing of $H_\phi$ all variable vertices are placed on a horizontal line and the clause vertices connect to the adjacent variable vertices either from above or from below the horizontal line.
	In our gadget-based reduction we mimic the drawing of $H_\phi$ by creating a 3-plane drawing $\Gamma_\phi$ as a set of line segments of uniform length and a value $L$ such that $\Gamma_\phi$ has a SPED with ink at least $L$ if and only if $\phi$ is satisfiable.
	Figure~\ref{fig:gadget_connection} shows a sketch of the gadgets.
%		\end{cutout}

@@

	Each variable gadget is a cycle of segment pairs, with at least one pair for each occurrence of the variable in $\phi$, see Fig.~\ref{fig:gadget_connection}. 
	Observe that this cycle has exactly two ink-maximal SPEDs: either all red edges are full segments and all green edges are length-1 stubs or vice versa.
	We associate the configuration with green stubs and full red segments with the value \emph{true} and the configuration with full green segments and red stubs with the value \emph{false}.

@@

Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. 
Now we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. In both cases we present polynomial-time dynamic programming algorithms. Let $ C = (V,E) $ be the intersection graph of a given drawing $\Gamma$ of graph $ G $ as defined in Section~\ref{sec:preliminaries}. 
Let $ u \in V $ and $ \delta  = \deg(u) $. 
Then for the corresponding segment $ s(u) \in S $ there are $ \delta + 1 $ relevant stub pairs including the whole segment, see Fig.~\ref{fig:tree}. 
Let $ \ell_1(u), \dots, \ell_\delta(u) \in \mathbb{R}_+$ be the stub lengths induced by the intersection points of $s(u)$ with the segments of the neighbors of $u$,  sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole segment $ s(u) $.

@@

Recurrence~(\ref{rec:tree}) can be solved naively in $ O(mk^2) $ time, where $m = |V|$. 
Using the order on the stub lengths  we can improve this to $ O(mk) $ time by computing all $ T_i(u) $ for one $ u \in V $ in $O(k)$ time.
Let $u \in V$ be a vertex with degree $\deg(u) = \delta$. 
The values $T_0(u) = \ell_0(u) + \sum_{v\in c(u)} \solshort(v)$ and $T_1(u) = \ell_1(u) + \sum_{v\in c(u)} \sollong(v)$ for the whole segment $s(u)$ and the shortest stubs can be computed in $O(k)$ time each.
Now $T_{j+1}(u)$ can be computed from $T_j(u)$ in $O(1)$ time as follows.
Let $v_j $ be the neighbor of $u$ that induces stub length $\ell_j(u)$ and assume $v_j \ne p(u)$.
In $T_j(u)$ we could still count the value $\sollong(v_j)$, but in $T_{j+1}(u)$ the stub length of $u$ implies that $v_j$ can contribute only $\solshort(v_j)$. 
Then $T_{j+1}(u) = T_j(u) - \sollong(v_j) + \solshort(v_j)$.
If $v_j=p(u)$ the two values $T_j(u)$ and $T_{j+1}(u)$ are equal as the corresponding change in stub length has no effect on the children of $u$.
Computing $\solshort(u)$ and $\sollong(u)$ clearly takes $O(k)$ time.

@@

\begin{theorem}
	\label{thm:tree}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with maximum degree $ k \in \mathbb{N} $, problem \maxsped can be solved in $ O(mk + m \log m) $ time and space.
\end{theorem}
% \begin{proof}
% 	Since $ C $ is known to be a tree, and hence $ |E| = O(m) $, we can compute the intersection graph and the stub lengths in $ O(m\log m) $ time and space.
% 	With Lemma~\ref{lem:compute_vertex} and Recurrence~\ref{rec:tree} we need one pass through the tree to find the roots maximum in $ O(mk) $ time. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.
% %	To solve the \maxsped problem on $ G $ we need to compute $ \max \{T_i(r) \mid 0 \leq i \leq k \} $ for the root $ r \in V $ of $ C $ as seen above. Doing this bottom up requires one pass through $ C $ hence by applying Lemma~\ref{lem:compute_vertex} we need $ O(mk) $ time to compute all $ T_i(u) $ for all $ u \in V $. For every vertex we only need to compute $ O(k) $ values and store two of them, namely $ \sollong(u) $ and $ \solshort(u) $, so we do not need more than $ O(mk) $ memory. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.
% %
%
%
%
% 	Retrieving the actual stub length for each edge in $ G $ needed to produce the SPED can be done by backtracking the subsolutions which requires at most $ O(mk) $ time.
% \end{proof}

@@

T_{0}(v) & \text{otherwise}

@@

>>>>

The idea of drawing edges only partially has been formalized in graph drawing as follows~\cite{bk-ecbe-12}. 
A \emph{partial edge drawing (PED)} is a graph drawing that maps vertices to points and edges to pairs of crossing-free edge stubs of positive length pointing towards each other.
These edge stubs are obtained by erasing one contiguous central piece of the straight-line segment connecting the two endpoints of each edge.
In other words each straight-line edge is divided into three parts, of which only the two outer ones are drawn (see Fig.~\ref{fig:examples}).
More restricted and better readable~\cite{blmt-pedhmitc-16} variations of PEDs are \emph{symmetric} PEDs, in which both stubs of an edge must have the same length (see Fig.~\ref{fig:examples}(b)), and \emph{homogeneous} PEDs, in which the ratio of the stub length to the total edge length is the same for all edges.
\textcolor{blue}{Symmetric stubs facilitate finding adjacent vertices due to the identical stub lengths at both vertices, and symmetric homogeneous stubs additionally indicate the distance at which to find a neighboring vertex.}
The natural optimization problem in this formal setting is \emph{ink maximization}, i.e., maximizing the total stub length, so that as much information as possible is given in the drawing while all crossings disappear in the negative background space.

@@

Throughout the paper let $ G $ be a \emph{simple graph} with edge set $ S = \{s_1,\dots,s_m\}$ and $ \Gamma $ a straight-line drawing of $ G $ in the plane. We call $ \Gamma $ \emph{$ k $-plane} if every edge $ s_i \in S $ is crossed by at most $ k $ other edges from $ S $ in $ \Gamma $. We use the terms edge in $ S $ and segment in $ \Gamma $ interchangeably. %, since our input is always a graph $G$ with a given drawing $ \Gamma $. 
Hence we can also interpret $S$ as a set of line segments.

@@

%\renewcommand\windowpagestuff{%
%	\includegraphics{Reduction_visual}
%	\captionof{figure}{Example of three variable gadgets and\\ a satisfied clause gadget. Dotted parts do not belong\\ to the SPED.}
%	\label{fig:gadget_connection}
%}
%\opencutleft
\begin{proof}
%		\begin{cutout}{2}{0pt}{0.44\textwidth}{15}
	We reduce from the \NP-hard problem \ppsat~\cite{l-pftu-82} using similar ideas as in Bruckdorfer's sketch of the hardness proof for general \maxsped~\cite{b-sgh-15}. 
%	Bruckdorfer reduces from the more restricted problem \textsc{Planar Positive 1-in-3 Sat}.
	Let $\phi$ be a planar 3-Sat formula with $n$ variables $\{x_1, \dots, x_n\}$ and $m$ clauses $\{c_1, \dots, c_m\}$, each consisting of three literals.
	We can assume that $\phi$ comes with a planar drawing of its variable-clause graph $H_\phi$, which has a vertex for each variable $x_i$ and a vertex for each clause $c_j$.
	Each clause vertex is connected to the three variables appearing in the clause.
	In the drawing of $H_\phi$ all variable vertices are placed on a horizontal line and the clause vertices connect to the adjacent variable vertices either from above or from below the horizontal line.
	In our gadget-based reduction we mimic the drawing of $H_\phi$ by creating a 3-plane drawing $\Gamma_\phi$ as a set of line segments of uniform length and a value $L$ such that $\Gamma_\phi$ has a SPED with ink at least $L$ if and only if $\phi$ is satisfiable.
	Figure~\ref{fig:gadget_connection} shows a sketch of the gadgets.
%		\end{cutout}

@@

	Each variable gadget is a cycle of segment pairs, with (at least) one pair for each occurrence of the variable in $\phi$, see Fig.~\ref{fig:gadget_connection}. 
	Observe that this cycle has exactly two ink-maximal SPEDs: either all red edges are full segments and all green edges are length-1 stubs or vice versa.
	We associate the configuration with green stubs and full red segments with the value \emph{true} and the configuration with full green segments and red stubs with the value \emph{false}.

@@

Section~\ref{sec:hardness} showed that \maxsped is generally \NP-hard for $k \ge 3$. 
Now we consider the special case that the intersection graph of the $k$-plane input drawing is a tree or a cactus. In both cases we present polynomial-time dynamic programming algorithms. Let $ C = (V,E) $ be the intersection graph of a given drawing $\Gamma$ of a graph $ G $ as defined in Section~\ref{sec:preliminaries}. 
Let $ u \in V $ and $ \delta  = \deg(u) $. 
Then for the corresponding segment $ s(u) \in S $ there are $ \delta + 1 $ relevant stub pairs including the whole segment, see Fig.~\ref{fig:tree}. 
Let $ \ell_1(u), \dots, \ell_\delta(u) \in \mathbb{R}_+$ be the stub lengths induced by the intersection points of $s(u)$ with the segments of the neighbors of $u$,  sorted from shorter to longer stubs. We define  $ \ell_0(u) $ as the length of the whole segment $ s(u) $.

@@

Recurrence~(\ref{rec:tree}) can be solved naively in $ O(mk^2) $ time, where $m = |V|$. 
Using the order on the stub lengths  we can improve this to $ O(mk) $ time by computing all $ T_i(u) $ for one $ u \in V $ in $O(k)$ time.
Let $u \in V$ be a vertex with degree $\deg(u) = \delta$. 
The values $T_0(u) = \ell_0(u) + \sum_{v\in c(u)} \solshort(v)$ and $T_1(u) = \ell_1(u) + \sum_{v\in c(u)} \sollong(v)$ for the whole segment $s(u)$ and the shortest stubs can be computed in $O(k)$ time each.
Now $T_{j+1}(u)$ can be computed from $T_j(u)$ in $O(1)$ time as follows.
Let $v_j $ be the neighbor of $u$ that induces stub length $\ell_j(u)$ and assume $v_j \ne p(u)$.
In $T_j(u)$ we could still count the value $\sollong(v_j)$, but in $T_{j+1}(u)$ the stub length of $u$ implies that $v_j$ can contribute only to $\solshort(v_j)$. 
Then $T_{j+1}(u) = T_j(u) - \sollong(v_j) + \solshort(v_j)$.
If $v_j=p(u)$, then the two values $T_j(u)$ and $T_{j+1}(u)$ are equal as the corresponding change in stub length has no effect on the children of $u$.
Computing $\solshort(u)$ and $\sollong(u)$ clearly takes $O(k)$ time.

@@

\begin{theorem}
	\label{thm:tree}
	Let $ G $ be a simple graph with $m$ edges and $ \Gamma $ a straight-line drawing of $ G $. If the intersection graph $ C = (V,E) $ of $ G $ is a tree with maximum degree $ k \in \mathbb{N} $, then problem \maxsped can be solved in $ O(mk + m \log m) $ time and space.
\end{theorem}
% \begin{proof}
% 	Since $ C $ is known to be a tree, and hence $ |E| = O(m) $, we can compute the intersection graph and the stub lengths in $ O(m\log m) $ time and space.
% 	With Lemma~\ref{lem:compute_vertex} and Recurrence~\ref{rec:tree} we need one pass through the tree to find the roots maximum in $ O(mk) $ time. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.
% %	To solve the \maxsped problem on $ G $ we need to compute $ \max \{T_i(r) \mid 0 \leq i \leq k \} $ for the root $ r \in V $ of $ C $ as seen above. Doing this bottom up requires one pass through $ C $ hence by applying Lemma~\ref{lem:compute_vertex} we need $ O(mk) $ time to compute all $ T_i(u) $ for all $ u \in V $. For every vertex we only need to compute $ O(k) $ values and store two of them, namely $ \sollong(u) $ and $ \solshort(u) $, so we do not need more than $ O(mk) $ memory. Adding the bounds yields the desired runtime of $ O(mk + m \log m)) $.
% %
%
%
%
% 	Retrieving the actual stub length for each edge in $ G $ needed to produce the SPED can be done by backtracking the subsolutions which requires at most $ O(mk) $ time.
% \end{proof}

@@

T_{0}(v) & \text{otherwise.}

@@

\end{cases}
